Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2016-09-25 writeup/local_pca_paper.tex
grr submission rules

% % stuff for submission
% \usepackage[nomarkers,figuresonly]{endfloat}
% \usepackage{setspace}
% \doublespacing
% \usepackage{lineno}
% \linenumbers
% \bibliographystyle{plos2015}
% \newcommand\citet{\cite}
% \newcommand\citep{\cite}

% % for the nice version
\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\bibliographystyle{plainnat}

@@

\citet{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
where ``subdivision'' refers to restricted migration between subpopulations.
The phrase is also commonly used to refer to the genetic patterns that result from this process,
as for instance reduced mean relatedness between individuals from distinct populations.
However, it is not necessarily clear what aspects of demography should be included in the concept.
For instance, 
\citet{blair1943population} defines \emph{population structure} to be the sum total of
``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
The definition is similar to Wright's, but differs in including the effects of natural selection.
On closer examination, incorporating differential survival or fecundity makes the concept less clear:
should a randomly mating population consisting of two types that are partially reproductively isolated from each other
be said to show population structure or not?
Whatever the definition,
it is clear that due to natural selection, the effects of population structure --
the \emph{realized} patterns of genetic relatedness --
differ depending on which portion of the genome is being considered.
For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
increasing genetic differentiation between populations near to this gene.
Similarly, newly adaptive alleles spread first in local populations.
These observations motivate many methods to search for genetic loci under selection,
as for example in \citet{huertasnchez2013genetic} and \citet{martin2016natural}.

@@

Patterns in genome-wide relatedness are often summarized
by applying principal components analysis (PCA, \citep{patterson2006population}) 
to the genetic covariance matrix,
as pioneered by \citet{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
The results of PCA can be related to the genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
and sometimes produce ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes}.
% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

@@

% Fig fig:pca_by_pop WAS HERe

@@

>>>>

\newif\ifsubmission
% \submissiontrue
\submissionfalse
\ifsubmission
    % % stuff for submission
    % \usepackage[nomarkers,figuresonly]{endfloat}
    \usepackage{setspace}
    \doublespacing
    \usepackage{lineno}
    \linenumbers
    \bibliographystyle{plos2015}
    \newcommand\citet{\cite}
    \newcommand\citep{\cite}
    \renewcommand\includegraphics{\relax}
    \renewcommand{\includegraphics}[2][]{\relax}
\else
    % for the nice version
    \usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
    \bibliographystyle{plainnat}
\fi
% grrr
\newcommand{\citetx}[2][]{\ifsubmission#1 \cite{#2}\else\citet{#2}\fi}

@@

\citetx[Wright]{wright1949genetical} defined \emph{population structure} to encompass ``such matters as numbers, composition by age and sex, and state of subdivision'',
where ``subdivision'' refers to restricted migration between subpopulations.
The phrase is also commonly used to refer to the genetic patterns that result from this process,
as for instance reduced mean relatedness between individuals from distinct populations.
However, it is not necessarily clear what aspects of demography should be included in the concept.
For instance, 
\citetx[Blair]{blair1943population} defines \emph{population structure} to be the sum total of
``such factors as size of breeding populations, periodic fluctuation of population size, sex ratio, activity range and \emph{differential survival of progeny}'' (emphasis added).
The definition is similar to Wright's, but differs in including the effects of natural selection.
On closer examination, incorporating differential survival or fecundity makes the concept less clear:
should a randomly mating population consisting of two types that are partially reproductively isolated from each other
be said to show population structure or not?
Whatever the definition,
it is clear that due to natural selection, the effects of population structure --
the \emph{realized} patterns of genetic relatedness --
differ depending on which portion of the genome is being considered.
For instance, strongly locally adapted alleles of a gene will be selected against in migrants to different habitats,
increasing genetic differentiation between populations near to this gene.
Similarly, newly adaptive alleles spread first in local populations.
These observations motivate many methods to search for genetic loci under selection,
as for example in \citet{huertasnchez2013genetic} and \citet{martin2016natural}.

@@

Patterns in genome-wide relatedness are often summarized
by applying principal components analysis (PCA, \citep{patterson2006population}) 
to the genetic covariance matrix,
as pioneered by \citetx[Menozzi et al]{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
The results of PCA can be related to the genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
and sometimes produce ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes}.
% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

@@

@@

\section{Supplementary Figures}
